Ornaments as indicators of social changes in northeastern Taiwan before and after the European colonial period

### 2020-05-14 analysis/paper/paper.Rmd
editing for re-submission after peer reviews

This article describes personal ornaments excavated from the upper component of Kiwulan (1350 AD-1850 AD), the largest Iron Age settlement on the Yilan plain in northeastern Taiwan. The first recorded European presence in Yilan was a Spanish revenge attack on Indigenous villages in 1632 [@Borao2001, pp. 163]. In 1647 the Dutch attacked villages and forced them to accept colonial rules and pay an annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest Indigenous settlement in the plain, with a population of 840 adults [@Nakamura1938, pp. 12]. Following defeat of Dutch by the Chinese general Koxinga in 1661-1662, the Dutch abandoned northern Taiwan. Direct contact with Han Chinese is indicated by Qing dynasty census reports mentioning Yilan villages in 1821 [@Yao1996].

@@

The island of Taiwan lies at the junction of mainland China, Southeast Asia, and Northeast Asia in the Pacific Ocean. The prehistory of Taiwan island could be roughly divided into three major periods, palaeolithic (c. 27,000 BP- 5000BP), Neolithic (c. 6500- 2000BP), and Iron age (c. 2000- 400BP) with slightly regional differences in onset of each period and variations in style of artifacts and assemblages [@Chen2017; @Liu2011]. It is generally accepted that Taiwan entered the historical period in the early 17th century due to the colonial activities of the Spanish and the Dutch who played an important role in keeping written records about Taiwan. The European colonial presence in Taiwan ended in 1662 when the Dutch were defeated by the kingdom of Tungning, founded by Koxinga from China. Later in 1683, Taiwan was incorporated into the Qing dynasty in China and a large wave of Han Chinese migrated to Taiwan during the late 18th century. Because of natural safe harbors, northeastern Taiwan was involved in a regional trade network through cross-culture interactions with Chinese merchants since the 14th century, and later the global trade network with the Europeans in the 17th century brought more trade goods circulated in Southeast Asia into Taiwan [@Chen2005; @Liu2017]. Although located on the periphery of regional trade centers, Yilan was connected to trade networks via visits of other Indigenous groups, Chinese merchants, and Europeans, via sea.

The European presence in northern Taiwan started with the Spanish who founded Fort San Salvador at Helping dao, Keelung in 1626, and Fort San Domingo in 1629 at Tamsui (Figure \@ref(fig:KWL-map)). They sent missionaries to local Indigenous settlements in this region [@Blusse2000, pp. 343] and kept records about their observations of Indigenous communities. A Dominican priest in 1632 reported that the Taparri, an Indigenous tribe from northern Taiwan, exchanged carnelian beads with other Indigenous groups. This form of exchange was widespread and even the Spanish soldiers used carnelian beads as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is further indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the female shamans in the tribe would use carnelian beads as magical items in ritual healing practices [@Borao2009, pp. 122-151]. Records of an Indigenous funeral document the use of carnelian beads in ritual contexts, with more carnelian beads, pottery, and cloth placed into the graves of more influential people to indicate their family's higher status [@LiandWu2006, pp. 153]. While a full critical analysis of these historical accounts remains to be produced, we take them to minimally indicate that carnelian beads were already treated as prestige goods in Yilan before the arrival of Europeans. In 1642, the Dutch Vereenigde Oostindische Compagnie (VOC) defeated the Spanish and took over their forts in northern Taiwan. They introduced a feudal system in an attempt to control the Indigenous communities by asking Indigenous leaders to attend an annual ceremony for demonstrating their loyalty and paying tribute [@Andrade2007, ch. 9; @Kang2016, ch. 4]. The Dutch provided beads and other goods based on negotiations with Indigenous communities to secure alliances in the annual ceremony or during their travels [@Kang2016, ch. 6]. We might predict that the activities of the Dutch feudal system to build and maintain alliances resulted in an increase in the amount and diversity of ornaments in northeastern Indigenous communities during this period.

@@

Information about Kiwulan (Figure \@ref(fig:KWL-map)) comes from a rescue archaeology project that was conducted between 2001-2004 in advance of a water diversion project and road bridge construction. The excavations used 2 mm and 1.5 mm mesh screens and covered eight open area sections in total of 262 squares (4 m by 4 m) reaching 3,814 m^2^ [@Chen2007]. The nearly 2 m thick archaeological deposits reveal a large amount of artifacts, burials, middens, post-holes, wooden pillars, and stone structures, all of which indicates it was a long-term settlement. Artifact locations were recorded to the 2 x 2 m sub-square they were recovered in; they lack individual point provenance. Based on the continuity of deposition and the frequency of artifacts, the center of the site is the open area consisting of A and D sections, which is also the study area where our samples selected from (Figure \@ref(fig:KWL-sam-area)). In the AD area, post-holes were found aligned in a north-south direction in some intervals with construction marks, which were interpreted as the remains of silt house structures. At the north margin of the dwelling place were burials that are mostly oriented in an east-west direction.

@@

```{r, KWL-sam-area, fig.cap = "Map showing the largest section of excavation areas at Kiwulan, and the distribution of forty squares sampled in this paper presented in red with square ID number. Small dots represent the location of post-holes. Each square is 4 x 4 m", fig.width=image_width_2_col_inch}
# add to the map and scale and north arrow. Check it out after knitting 
library(sf)
library(cartography)
#devtools::install_github('oswaldosantos/ggsn')
library(ggsn)
# read in the spatial data
burial <- invisible(st_read(here("analysis", "data", "raw_data", "AD_burial.shp"), quiet = TRUE))
post <- invisible(st_read(here("analysis", "data", "raw_data", "AD_postholes.shp"), quiet = TRUE))
AD <- invisible(st_read(here("analysis", "data", "raw_data", "AD_zone.shp"), quiet = TRUE))
AD_data <- invisible(st_read(here("analysis", "data", "raw_data", "AD_withdata.shp"), quiet = TRUE))
location <- invisible(st_read(here("analysis", "data", "raw_data", "location.shp"), quiet = TRUE))
units_40 <- invisible(read.csv(here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv")))

@@

The chronology of Kiwulan can be divided into two phases represented by a upper component (1350-1850 AD, 600-100 BP) and a lower component (650-1150 AD, 1200-800 BP) separated by a sterile deposit spanning c. 150 years [@Chen2007]. The sterility may be related to dry weather leading to site abandonment according to pollen analysis [@Lin2015]. These component divisions are based on the differences in color and texture of the deposit and the content of artifacts. Whether these two components represent a culture continuity from the same people is still under debate [@Chiu2004; @Chen2007]. There are 32 radiocarbon ages spanning the two components, previously published by @Chen2007, and shown here in Figure \@ref(fig:C14-data) and Table \@ref(tab:C14-table-data). We focus on the upper component because only this component spans the periods of pre-European contact, European presence, and Chinese presence. In the upper component, all excavation squares in our sampling area show signs of continuous human occupation during each of the three phases. Previous work divided the upper component into six analytical units, spanning from the 14th century to the 19th century, according to the radiocarbon dates, excavation depth, consistency of contexts, and types of chronologically diagnostic ceramics such as blue and white porcelains [@Hsieh2009; @Wang2011]. Based on these six units and copies of original excavation records, we have made some refinements of the original chronology to assign contexts relative to the pre-European, European, and the Chinese periods. The methods we used to devise the new chronology are described in the Supplementary Information document.

The archaeological indicators of the start of the European influence at Kiwulan are the appearance of light grey glazed jars, known as "An-ping" jars in China and Taiwan, and large dark brown glazed stoneware jars that were introduced to Taiwan during the early 17th century. Large dark brown glazed stoneware jars may have been made in Southeast Asia, and are frequently found in European shipwrecks from this period as vessels for transporting water, wine or other liquids on long voyages. The earliest evidence of light gray glazed jars in this region has been found among the cargo of the Spanish shipwreck _San Diego_, which sunk in 1600 AD [@Dizon2016; @Hsieh1995]. Southeast China is assumed to be the origin of the light gray glazed jars, however these are commonly found at sites in Taiwan that were associated with European activities, such as the Zeelandia fort site in Tainan [@Wang2007]. The jar shapes found at Kiwulan are typical of those found elsewhere in VOC sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. We cannot be sure of the exact process that brought them to Kiwulan: they might have been directly imported by Europeans, by Chinese merchants, or by Indigenous groups via regional networks in north Taiwan. In any case, the high volume of ceramics transported by Europeans, and their high mobility in shipping trade played an important role in introducing foreign jars to Taiwan.

Those jars were widely distributed across the site and can serve as indicators, together with the radiocarbon dates, to identify the excavation units associated with the pre-European period and the start of European influence at Kiwulan. In addition to stoneware jars as indicators of European presence, around 300 pieces of locally made clay pipes and a few imported pipes were found at Kiwulan. Smoking is likely to have been introduced by Europeans. This custom was widely adopted in many European countries in the 16th century and spread to other regions through interactions [@Uccar2019]. We found that the presence of pipe bowls in the archaeological record here is consistent with distributions of glazed jar fragments, which are far more numerous and widespread across the site (n = 1685). It should be noted that the temporal indicators might have been introduced before the European arrival by Chinese traders that could mislead the interpretation of the start of European phase. However, the archaeological evidence shows that the layers with abundant trade ceramics match the 17th century according to excavation report and previous studies [@Hsieh2009; @Wang2011]. Thus, we focus on identifying the context with high frequency of those ceramics indicating the early 17th century when they prevailed.

The archaeological signature of the Chinese period at Kiwulan is the large amount and diversity of Chinese porcelains in many styles and forms such as bowls, plates, and cups. Other indicators include opium pipe-bowls and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Chinese migrations to Yilan were also recorded in official Chinese records written in the early 19th century recording the first immigrants in 1768 [@Chen1963; @Ke1993]. The details about the distribution of the temporal indicators, accompanied with stratigraphy data, radiocarbon dates, and archaeological contexts for each sampled excavation unit are provided in Supplementary Information data.

@@

Ornaments (Figure \@ref(fig:ornaments-image)) were found in a variety of archaeological contexts including post-holes area, burials, and middens. For the ornament categories, we follow the well established topology based on raw materials and shapes for the region of northeastern Taiwan [@Chen2007; @NMTH2005]. This study focuses on `r sam_post_hole` ornaments from 40 sampling squares located at the dwelling place of Kiwulan, indicated by aligned post-holes with _in-situ_ posts (Figure \@ref(fig:KWL-sam-area)). Occupation floors were not identified during excavation. We choose these units because they were stratigraphically intact and undisturbed by modern construction activity, compared to excavation squares on the periphery of the site. There are 35 burials in the sampling area, one third of the total number of burials at Kiwulan. Ornaments are commonly used as grave goods in burials, with the total number of ornaments in burials including 3,173. Besides, there are 27 ornaments found in midden contexts. The obvious difference in number in burials and midden contexts indicates the property of ornaments as prestige goods. The high number in burials is due to the presence of bead strands or patterned bands of beads, which sometimes contains thousands of beads in an individual burial [@Chen2007]. The unequal distribution of beads between burials was interpreted as the evidence for hierarchy by @Cheng2008. However, @Hsieh2012 suggests a relatively egalitarian form from the comparative analysis of frequency and proportion of burial goods. To have a comprehensive understanding, we focus on ornaments from the dwelling place contexts (Figure \@ref(fig:plot-all-ornaments-count), Table \@ref(tab:tidy-data-for-table)) because these give us the greatest spatial and temporal representation across the three time periods, and so are most informative of social inequality as indicated by uneven distributions of ornaments. The burials are excluded because most burials from the sampling area date to the European period (n = 21), limiting the usefulness of comparisons between the periods. Ornaments could be potential heirlooms inherited over multiple generations and preserved well for a long time. However, there is no continuous increase over time that suggests accumulation and discard of ornaments is not constant. Thus, we assume that changes in the abundance of ornaments reflect changes in the amount in archaeological record.

```{r, ornaments-image, fig.cap = "Subtypes of ornament in each major class. A: carnelian beads, B: glass beads, C: gold-foil beads, D: bells, E: metal rings. Photographs are presented in the same order as those subtypes in the table but from left to right instead. The photographs of B, C, D, E classes were from original excavation report (Chen 2007)", echo=FALSE}
knitr::include_graphics(here("analysis", "figures", "ornament_edit.png"))
```

@@

To enable re-use of materials and improve reproducibility and transparency [@Marwick2017], the entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in http://doi.org/10.17605/OSF.IO/R8YGA. Also in this version-controlled compendium [@Marwick2018] are the raw data for all the visualizations and tests reported here. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

@@

```{r, plot-counts-by-period, fig.cap = "Frequency of the major ornament across different time periods"}
### delete one plot
# categories of ornaments by period
ornaments_count <-
  ornaments_period_clean %>% 
  filter(!is.na(`6-layers`),
         Categories != "Unknown Metal") %>%
  ungroup() %>% 
  mutate(Categories = fct_lump(Categories, 
                               n = 5)) %>% 
  mutate(Categories = fct_infreq(Categories)) %>% 
  mutate(Categories = fct_rev(Categories)) %>% 
  group_by(period, Categories) %>%
  count()
# histogram facetting for three periods
plot_ornaments_count_three_periods <- 
ggplot(ornaments_count) +
  geom_col(aes(Categories, n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(x = "Categories",
       y = "Frequency") +
  theme_minimal(base_size = 10)

@@

Figure \@ref(fig:plot-counts-by-period) shows the comparison of frequencies of the major classes of ornaments for different time periods at Kiwulan. The difference in frequencies between the three time periods reflect significant differences in the use of ornaments (chi-square = `r chisq_test_result_stat`, df = `r chisq_test_result_df`, p-value = $`r chisq_test_result_p_value`$). Most ornament types were present before European contact. Ornament frequencies reached a peak during the European period and then dropped during the Chinese period, especially gold-foil beads. This trend can be also seen on other ornaments including carnelian beads, metal rings, and bells. However, glass beads show a different pattern that indicates a higher frequency in the pre-European contact, and then a decrease in the European period and a further decrease in the Chinese period. To model the number of ornaments as a function of the mass of ceramics in each period, a Poisson GLM with a log link function was used. The model reveals that ceramic abundance strongly predicts the number of ornaments (β = $`r beta_print`$, p = $`r pval_print`$). If ceramic abundance is a suitable proxy for population at Kiwulan due to its basic role as cooking vessels, then ornament quantities per period may be influenced by the number of people living at the site.

@@

```{r plot-altogether, fig.cap= "Frequency of ornament subtypes showing the changes in frequency across time periods for metal rings, carnelian beads, bells, and glass beads"}
library(cowplot)
plot_multi_types_by_period <- 
plot_grid(ring_shape_plot,
          agate_bead_shape_plot, 
          bell_shape_plot, 
          glass_bead_plot,
          ncol=1,
          align = "h",
          axis = "lr",
          rel_widths = 1,
          rel_heights = c(1,1,0.7,0.62))
ggsave(plot = plot_multi_types_by_period,
       here("analysis", "figures",
       "plot-multi-types-by-period.jpg"),
       width = 135, 
       dpi = 300,
       height = 95,
       units = "mm")
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-multi-types-by-period.jpg"))
```

@@

The distribution of frequencies for subtypes in each major class is presented in Figure \@ref(fig:plot-altogether). Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture interaction instead of the effects of sample size. Carnelian beads and metal rings have greater quantity and variety of shapes compared to copper bells and glass beads during the European period. The greater varieties for carnelian beads and metal rings might indicate multiple origins due to participation in large scale trade networks stimulated by the European presence. In contrast, copper bells have less variety typically >2 cm long with a wide variety of human faces as a motif. Although glass beads have less variety in size, small (0.5-1 cm) and medium (1-2 cm), they have a wide variety of colors or patterns mostly made by a winding technique and with high lead content indicating possibly from China [@Cheng2008]. Although we are not certain of the specific origin of these beads, research suggest that these glass beads and metal ornaments have similar production techniques and composition to those found in China [@Chen2011; @Wang2018]. There seem to be no obvious changes in the sources of glass beads or metal ornaments at different periods in the upper component of Kiwulan (1350-1850 AD). However, the glass beads found from the lower component (650-1150 AD) are mostly Indo-Pacific beads, widespread in Southeast Asian sites from 300 BC until the early 2nd millennium [@Francis2002; @Wang2018].

@@

```{r plot-all-spatial-pattern, fig.cap= "Spatial densities of all class of ornament by time periods"}
library(sf)
library(viridis)

@@

```{r, plot-spatial-pattern-each, fig.cap="Spatial densities for ornament class by time periods, only those types with more than 5 pieces are shown here"}
ornaments_period_join_selected_types <- 
  ornaments_period_join %>% 
  filter(Categories %in% c('Gold-foil bead', 
                           'Glass bead',
                           'Carnelian bead'))

@@

```{r hypo-test-distribution, fig.cap= "Histograms of simulated ANN values from 1000 simulations for three time periods. X-axis values represent ANN expected value under a completely random process resulted from a simulated pattern. Each sample distribution presents the null hypothesis with the blue line indicating the observed ANN value"}

@@

##### pagebreak

>>>>

This article describes personal ornaments excavated from the upper component of Kiwulan (1350 AD-1850 AD), the largest Iron Age settlement on the Yilan plain in northeastern Taiwan. Ornaments are found at many Iron Age sites in Yilan, but only Kiwulan shows clear stratigraphic contexts from pre-European period to modern time. The first recorded European presence in Yilan was a Spanish revenge attack on Indigenous villages in 1632 [@Borao2001, pp. 163]. In 1647 the Dutch attacked villages and forced them to accept colonial rule and pay an annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest Indigenous settlement in the plain, with a population of 840 adults [@Nakamura1938, pp. 12]. Following defeat of Dutch by the Chinese general Koxinga in 1661-1662, the Dutch abandoned northern Taiwan. Direct contact with Han Chinese is indicated by Qing dynasty census reports mentioning Yilan villages in 1821 [@Yao1996].

@@

The island of Taiwan lies at the junction of mainland China, Southeast Asia, and Northeast Asia in the Pacific Ocean. The prehistory of Taiwan island could be roughly divided into three major periods, Palaeolithic (c. 27,000 BP- 5000BP), Neolithic (c. 6500- 2000BP), and Iron age (c. 2000- 400BP) with slight regional differences in onset of each period and variations in style of artifacts and assemblages [@Chen2017; @Liu2011]. It is generally accepted that Taiwan entered the historical period in the early 17th century due to the colonial activities of the Spanish and the Dutch who played an important role in keeping written records about Taiwan. The European colonial presence in Taiwan ended in 1662 when the Dutch were defeated by the kingdom of Tungning, founded by Koxinga from China. Later in 1683, Taiwan was incorporated into the Qing dynasty in China and a large wave of Han Chinese migrated to Taiwan during the late 18th century. Because of natural safe harbors, northeastern Taiwan was involved in a regional trade network through cross-culture interactions with Chinese merchants since the 14th century, and later the global trade network with the Europeans in the 17th century brought more trade goods circulated in Southeast Asia into Taiwan [@Chen2005; @Liu2017]. Although located on the periphery of regional trade centers, Yilan was connected to trade networks via visits of other Indigenous groups, Chinese merchants, and Europeans, via sea.

The European presence in northern Taiwan started with the Spanish who founded Fort San Salvador at Heping Dao, Keelung in 1626, and Fort San Domingo in 1629 at Tamsui (Figure \@ref(fig:KWL-map)). They sent missionaries to local Indigenous settlements in this region [@Blusse2000, pp. 343] and kept records about their observations of Indigenous communities. A Dominican priest in 1632 reported that the Taparri, an Indigenous tribe from northern Taiwan, exchanged carnelian beads with other Indigenous groups. This form of exchange was widespread and even the Spanish soldiers used carnelian beads as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is further indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the women shamans in the tribe would use carnelian beads as magical items in ritual healing practices [@Borao2009, pp. 122-151]. Records of an Indigenous funeral describe the use of carnelian beads in ritual contexts, with more carnelian beads, pottery, and cloth placed into the graves of more influential people to indicate their family's higher status [@LiandWu2006, pp. 153]. While a full critical analysis of these historical accounts remains to be produced, we take them to minimally indicate that carnelian beads were already treated as prestige goods in Yilan before the arrival of Europeans. In 1642, the Dutch Vereenigde Oostindische Compagnie (VOC) defeated the Spanish and took over their forts in northern Taiwan. They introduced a feudal system in an attempt to control the Indigenous communities by asking Indigenous leaders to attend an annual ceremony for demonstrating their loyalty and paying tribute [@Andrade2007, ch. 9; @Kang2016, ch. 4]. The Dutch provided beads and other goods based on negotiations with Indigenous communities to secure alliances in the annual ceremony or during their travels [@Kang2016, ch. 6]. We might predict that the activities of the Dutch feudal system to build and maintain alliances resulted in an increase in the amount and diversity of ornaments in northeastern Indigenous communities during this period.

@@

Archaeological ornaments from Kiwulan (Figure \@ref(fig:KWL-map)) come from a rescue archaeology project that was conducted between 2001-2004 in advance of a water diversion project and road bridge construction. The excavations used 2 mm and 1.5 mm mesh screens and covered eight open area sections in total of 262 squares (4 m by 4 m) reaching 3,814 m^2^ [@Chen2007]. The nearly 2 m thick archaeological deposits reveal a large amount of artifacts, burials, middens, post-holes, wooden pillars, and stone structures, all of which indicates it was a long-term settlement. Artifact locations were recorded to the 2 x 2 m sub-square they were recovered in; they lack individual point provenance. Based on the continuity of deposition and the frequency of artifacts, the center of the site is the open area consisting of the A and D sections, which is also the study area where our samples come from (Figure \@ref(fig:KWL-sam-area)). In the AD area, post-holes were found aligned in a north-south direction in intervals with construction marks, which have been interpreted as the remains of stilt house structures. At the north margin of the dwelling place were burials that are mostly oriented in an east-west direction.

@@

```{r, KWL-sam-area, fig.cap = "Map showing the largest section of excavation areas at Kiwulan, and the distribution of forty squares sampled in this paper presented in red with square ID numbers. Small dots represent the locations of post-holes. Each square is 4 x 4 m", fig.width=image_width_2_col_inch}
# add to the map and scale and north arrow. Check it out after knitting 
library(sf)
library(cartography)
#devtools::install_github('oswaldosantos/ggsn')
library(ggsn)
# read in the spatial data
burial <- invisible(st_read(here("analysis", "data", "raw_data", "AD_burial.shp"), quiet = TRUE))
post <- invisible(st_read(here("analysis", "data", "raw_data", "AD_postholes.shp"), quiet = TRUE))
AD <- invisible(st_read(here("analysis", "data", "raw_data", "AD_zone.shp"), quiet = TRUE))
AD_data <- invisible(st_read(here("analysis", "data", "raw_data", "AD_withdata.shp"), quiet = TRUE))
location <- invisible(st_read(here("analysis", "data", "raw_data", "location.shp"), quiet = TRUE))
units_40 <- invisible(read.csv(here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv")))

@@

The chronology of Kiwulan can be divided into two phases represented by a upper component (1350-1850 AD, 600-100 BP) and a lower component (650-1150 AD, 1200-800 BP) separated by a sterile deposit spanning c. 150 years [@Chen2007]. The sterility may be related to dry weather, according to pollen analysis, leading to site abandonment [@Lin2015]. These component divisions are based on the differences in color and texture of the deposit and the content of artifacts. Whether these two components represent a culture continuity from the same people is still under debate [@Chiu2004; @Chen2007]. There are 32 radiocarbon ages spanning the two components, previously published by @Chen2007, and shown here in Figure \@ref(fig:C14-data) and Table \@ref(tab:C14-table-data). We focus on the upper component because only this component spans the periods of pre-European contact, European presence, and Chinese presence. In the upper component, all excavation squares in our sampling area show signs of continuous human occupation during each of the three phases. Previous work divided the upper component into six analytical units, spanning from the 14th century to the 19th century, according to the radiocarbon dates, excavation depth, consistency of contexts, and types of chronologically diagnostic ceramics such as blue and white porcelains [@Hsieh2009; @Wang2011]. Based on these six units and copies of original excavation records, we have made some refinements of the original chronology to assign contexts into the pre-European, European, and the Chinese periods. The methods we used to devise the new chronology are described in the Supplementary Online Materials.

The archaeological indicators of the start of the European influence at Kiwulan are the appearance of light grey glazed jars, known as "An-ping" jars in China and Taiwan, and large dark brown glazed stoneware jars that were introduced to Taiwan during the early 17th century. Large dark brown glazed stoneware jars may have been made in Southeast Asia, but are frequently found in European shipwrecks from this period as vessels for transporting water, wine or other liquids on long voyages. The earliest evidence of light gray glazed jars in this region has been found among the cargo of the Spanish shipwreck _San Diego_, which sunk in 1600 AD [@Dizon2016; @Hsieh1995]. Southeast China is assumed to be the origin of the light gray glazed jars, however these are commonly found at sites in Taiwan that were associated with European activities, such as the Zeelandia fort site in Tainan [@Wang2007]. The jar shapes found at Kiwulan are typical of those found elsewhere in VOC sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. We cannot be sure of the exact process that brought them to Kiwulan: they might have been directly imported by Europeans, by Chinese merchants, or by Indigenous groups via regional networks in north Taiwan. In any case, the high volume of ceramics transported by Europeans, and their high mobility in the shipping trade played an important role in introducing foreign jars to Taiwan.

Those jars were widely distributed across the site and can serve as indicators, together with the radiocarbon dates, to identify the excavation units associated with the pre-European period and the start of European influence at Kiwulan. In addition to stoneware jars as indicators of European presence, around 300 pieces of locally made clay pipes and a few imported pipes were found at Kiwulan. Smoking is likely to have been introduced by Europeans. This custom was widely adopted in many European countries in the 16th century and spread to other regions through interactions [@Uccar2019]. We found that the presence of pipe bowls in the archaeological record here is consistent with distributions of glazed jar fragments, which are far more numerous and widespread across the site (n = 1685). It should be noted that these temporal indicators might have been introduced before direct European contact by Chinese traders, and this could result in some uncertainty in identifying the start of European phase. However, the archaeological evidence shows that the layers with abundant trade ceramics match the 17th century according to the excavation report and previous studies [@Hsieh2009; @Wang2011]. Thus, we focus on identifying the contexts with high frequencies of those ceramics as indicators the early 17th century.

The archaeological signature of the Chinese period at Kiwulan is the large amount and diversity of Chinese porcelains in many styles and forms such as bowls, plates, and cups. Other indicators include opium pipe-bowls and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Chinese migrations to Yilan were also recorded in official Chinese records written in the early 19th century recording the first immigrants in 1768 [@Chen1963; @Ke1993]. The details about the distribution of the temporal indicators, accompanied with stratigraphy data, radiocarbon dates, and archaeological contexts for each sampled excavation unit are provided in Supplementary Online Materials.

@@

Ornaments (Figure \@ref(fig:ornaments-image)) were found in a variety of archaeological contexts including post-holes area, burials, and middens. For the ornament categories, we follow the well-established topology based on raw materials and shapes for the region of northeastern Taiwan [@Chen2007; @NMTH2005]. This study focuses on `r sam_post_hole` ornaments from 40 sampling squares located at the main habitation areas of Kiwulan, indicated by aligned post-holes with _in-situ_ posts (Figure \@ref(fig:KWL-sam-area)). Occupation floors were not identified during excavation. We choose these units because they were stratigraphically intact and undisturbed by modern construction activity, compared to excavation squares on the periphery of the site. There are 35 burials in the sampling area, one third of the total number of burials at Kiwulan. Ornaments are commonly used as grave goods in burials, with the total number of ornaments in burials including 3,173. There are also 27 ornaments found in midden contexts. The obvious difference in number in burials and midden contexts indicates the property of ornaments as prestige goods. The high number in burials is due to the presence of bead strands or patterned bands of beads, which sometimes contain thousands of beads in an individual burial [@Chen2007]. The highly uneven distribution of beads between burials was interpreted as the evidence for hierarchy by @Cheng2008. However, @Hsieh2012 suggests a relatively egalitarian society from the comparative analysis of frequency and proportion of burial goods.

We focus on ornaments from the habitation contexts (Figure \@ref(fig:plot-all-ornaments-count), Table \@ref(tab:tidy-data-for-table)) because these give us the greatest spatial and temporal representation across the three time periods, and so are most informative of social inequality as indicated by uneven distributions of ornaments. The burials are excluded because most burials from the sampling area date to the European period (n = 21), limiting the usefulness of comparisons between the periods. A possible limitation to our chronological resolution is that ornaments could be heirlooms inherited over multiple generations and well-preserved for a long time. This is difficult to rule out completely, but we consider that because there is no continuous increase in ornament frequency over time, we conclude that accumulation and discard of ornaments is not constant. Thus, we assume that changes in the abundance of ornaments reflect relatively continuous discard behaviours rather than accumulations due to collecting of heirlooms.

```{r, ornaments-image, fig.cap = "Subtypes of ornament in each major class. A: carnelian beads, B: glass beads, C: gold-foil beads, D: bells, E: metal rings. Photographs are presented in the same order as those subtypes in the table but from left to right instead. The photographs of B, C, D, E classes are from the original excavation report (Chen 2007)", echo=FALSE}
knitr::include_graphics(here("analysis", "figures", "ornament_edit.png"))
```

@@

To enable re-use of materials and improve reproducibility and transparency [@Marwick2017], the entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at http://doi.org/10.17605/OSF.IO/R8YGA. Also in this version-controlled compendium [@Marwick2018] are the raw data for all the visualizations and tests reported here. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

@@

```{r, plot-counts-by-period, fig.cap = "Frequency of the major ornament types across different time periods."}
### delete one plot
# categories of ornaments by period
ornaments_count <-
  ornaments_period_clean %>% 
  filter(!is.na(`6-layers`),
         Categories != "Unknown Metal") %>%
  ungroup() %>% 
  mutate(Categories = fct_lump(Categories, 
                               n = 5)) %>% 
  mutate(Categories = fct_infreq(Categories)) %>% 
  mutate(Categories = fct_rev(Categories)) %>% 
  group_by(period, Categories) %>%
  count()
# histogram facetting for three periods
plot_ornaments_count_three_periods <- 
ggplot(ornaments_count) +
  geom_col(aes(Categories, n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(x = "Categories",
       y = "Frequency") +
  theme_minimal(base_size = 10)

@@

Figure \@ref(fig:plot-counts-by-period) shows the comparison of frequencies of the major classes of ornaments for different time periods at Kiwulan. The difference in frequencies between the three time periods reflect significant differences in the use of ornaments (chi-square = `r chisq_test_result_stat`, df = `r chisq_test_result_df`, p-value = $`r chisq_test_result_p_value`$). Most ornament types were present before European contact. Ornament frequencies reached a peak during the European period and then dropped during the Chinese period, especially gold-foil beads. This trend can be also seen on other ornaments including carnelian beads, metal rings, and bells. However, glass beads show a different pattern that indicates a higher frequency in the pre-European contact, and then a decrease in the European period and a further decrease in the Chinese period. To model the number of ornaments as a function of the mass of ceramics in each period, we used a Poisson GLM with a log link function. The model reveals that ceramic abundance strongly predicts the number of ornaments (β = $`r beta_print`$, p = $`r pval_print`$). If ceramic abundance is a suitable proxy for population at Kiwulan due to its basic role as cooking vessels, then ornament quantities per period may be influenced by the number of people living at the site.

@@

```{r plot-altogether, fig.cap= "Frequency of ornament subtypes showing the changes in frequency across time periods for metal rings, carnelian beads, bells, and glass beads."}
library(cowplot)
plot_multi_types_by_period <- 
plot_grid(ring_shape_plot,
          agate_bead_shape_plot, 
          bell_shape_plot, 
          glass_bead_plot,
          ncol=1,
          align = "h",
          axis = "lr",
          rel_widths = 1,
          rel_heights = c(1,1,0.7,0.62))
ggsave(plot = plot_multi_types_by_period,
       here("analysis", "figures",
       "plot-multi-types-by-period.jpg"),
       width = 135, 
       dpi = 300,
       height = 95,
       units = "mm")
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-multi-types-by-period.jpg"))
```

@@

Figure \@ref(fig:plot-altogether) shows the distribution of frequencies for subtypes in each major class. Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture interaction instead of the effects of sample size. Carnelian beads and metal rings have greater quantity and variety of shapes compared to copper bells and glass beads during the European period. The greater varieties for carnelian beads and metal rings might indicate multiple origins due to participation in large scale trade networks stimulated by the European presence. In contrast, copper bells have less variety, typically >2 cm long with a wide variety of human faces as a motif. Although glass beads have less variety in size, presenting as small (0.5-1 cm) or medium (1-2 cm), they have a wide variety of colors or patterns mostly made by a winding technique and with high lead content indicating possibly from China [@Cheng2008]. Although we are not certain of the specific origin of these beads, research suggest that these glass beads and metal ornaments have similar production techniques and composition to those found in China [@Chen2011; @Wang2018]. There seem to be no obvious changes in the sources of glass beads or metal ornaments at different periods in the upper component of Kiwulan (1350-1850 AD). However, the glass beads found from the lower component (650-1150 AD) are mostly Indo-Pacific beads, widespread in Southeast Asian sites from 300 BC until the early 2nd millennium [@Francis2002; @Wang2018].

@@

```{r plot-all-spatial-pattern, fig.cap= "Spatial densities of all class of ornament by time periods."}
library(sf)
library(viridis)

@@

```{r, plot-spatial-pattern-each, fig.cap="Spatial densities for ornament class by time periods, only those types with more than 5 pieces are shown here."}
ornaments_period_join_selected_types <- 
  ornaments_period_join %>% 
  filter(Categories %in% c('Gold-foil bead', 
                           'Glass bead',
                           'Carnelian bead'))

@@

```{r hypo-test-distribution, fig.cap= "Histograms of simulated ANN values from 1000 simulations for three time periods. X-axis values represent ANN expected values under a completely random process resulting from a simulated pattern. Each sample distribution presents the null hypothesis with the blue line indicating the observed ANN value"}

@@
